---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(environmentalhist)
library(stars); library(sf)
```

All raster data included has been stored as `stars` objects. See [this discussion on Github](https://github.com/r-spatial/stars/issues/633#issuecomment-1597356513) by Edzer Pebesma on the advantages and disadvantages of `stars` versus `terra` (the successor of `raster`) when it comes to working with raster data in R.

# Indices without a time component

To load the Ramankutty et al. (2002) index for agricultural suitability, simply type:

```{r}
data("ramankuttyetal2002")
```

This will store the raster as a `stars` object named `ramankutty` in your environment. It can be visualized by using the plot function:

```{r}
plot(ramankutty)
```

As you can see, these data come unprojected and the x-coordinates refer to degrees longitude and the y-coordinates to degrees latitude. The resolution is 0.5 degree and each cell value ranges between 0 and 1.


```{r}
sf::st_crs(ramankutty)
```



# Visualizing and using spatio-temporal data

To load the spatio-temporal index by Lehner, Philippe (2024) type:

```{r}
data("agrisuit_1500_2000")
```

The data come unprojected in EPSG:4326 with a WGS84 datum:

```{r}
sf::st_crs(suit1500200)
```

Each band of the raster refers to one year, starting from 1500:

```{r}
dim(suit1500200)
```


We can easily visualize every year separately. As per the vignette for `stars`, the subset operator `[` works as follows: 

* its first argument selects attributes
* second argument selects the first dimension
* third argument selects the second dimension, etc

To access the first year, we thus type:

```{r}
suit1500200[ , , ,1] |> 
  plot(main = "Agricultural Suitability in 1500")
```

We can visualize the temporal nature of the index using `ggplot2`:

```{r}
library(ggplot2)
library(ggthemes)

subset <- suit1500200[ , , ,c(1, 100, 200, 300)]
plot(subset)

ggplot() +
  geom_stars(data = subset) +
  facet_wrap("band") +
  scale_fill_viridis_c(option = "magma", name = "suitability") +
  coord_equal() +
  theme_map() +
  theme(legend.position = "bottom") +
  theme(legend.key.width = unit(2, "cm"))
```


# Manipulating rasters

For working with rasters, check the vignette of the `stars` package. We can cut, crop, or reproject the data as we please. As an example, we cut the global index to the bounding box of Europe:

```{r}
sf::st_crop(ramankutty, sf::st_bbox(suit1500200)) |> 
  plot(main = "Ramankutty et al. 2002 - Cropped to Europe")
```


# Transforming

You can of course also convert them to `terra` format if you don't feel comfortable working with `stars`:

```{r}
library(terra)
ramankutty.terra <- as(ramankutty, "SpatRaster")
plot(ramankutty.terra)

```

